---
title: "IPBLN Bioinformatics Report"

include-in-header:
  text: |
    <link rel="shortcut icon" href="../../../../../01-images/icons/favicon.svg"/>
    <link rel="icon" type="image/x-icon" href="../../../../../01-images/icons/favicon.ico">

params:
  project_path: ""
  experiment_name: ""
  report_language: ""
  report_version: ""
---


<!-- &&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&& -->
<!--  TAB 3: ANÁLISIS | 4. CUANTIFICACIÓN DE LA EXPRESIÓN GENÉTICA -->
<!-- &&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&& -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  CHATBOT -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

```{=html}
<!-- Contenedor del chat -->
<div id="chat-container">
  <button id="chat-toggle"></button>
  <div id="chat-box">
    <div id="chat-info" class="chat-info">
      <h1>
        Mini Chat RAG (beta)
      </h1>

      <p>
        Hello! I am <strong>Geni</strong>, the intelligent assistant of <strong>GenoScribe</strong>.  
        I am here to help you interactively explore the content of this bioinformatics report.
      </p>

      <p>
        When you ask me a question, I first try to recognize whether it matches any of the patterns or expressions I know.  
        If I find a match, I will respond directly with a predefined answer, designed to be fast, clear, and even a bit witty.  
        If I do not recognize the pattern, I then activate my search tools: I generate vector representations (embeddings) and look for the most relevant fragments across multiple documents—including the report itself, external PDF and HTML files, and question-and-answer (QA) sessions.  
        Based on that information, I create a summary that aims to provide you with a coherent and useful response grounded in the existing content.
      </p>

      <p>
        Please note that this environment is experimental. I do not use large language models, so some responses may be approximate or incomplete.  
        The main goal is to provide a fast, understandable, and reproducible way to visualize the information contained in the documents, enabling a more dynamic exploration of the report.
      </p>

      <p>
        Currently, results may vary in accuracy, as I use lightweight local models to ensure the application runs in any environment without the need for external servers.  
        However, the system structure is designed to significantly improve its performance in the future through integration with more advanced models or external APIs.  
        To get started, simply type your question in the field below and let me take care of the rest. I promise to put all my code into it!
      </p>
    </div>
    <div id="chat-log"></div>
    <input id="chat-input" type="text" placeholder="Escribe tu pregunta..."/>
    <button id="chat-send" title="Enviar">
      <svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 24 24">
        <path d="M2 21l21-9L2 3v7l15 2-15 2v7z"/>
      </svg>
    </button>
  </div>
</div>
```















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  IMPORTACIONES -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<!-- Incluir Font Awesome -->
<link href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/6.5.0/css/all.min.css" rel="stylesheet">















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  CÓDIGO GLOBAL -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

```{r}
library(glue)
library(fs)
library(stringr)
library(dplyr)
library(readr)
library(htmlwidgets)

# Rutas clave
ruta_qmd <- getwd()
ruta_bulk_rna_seq_pipeline <- file.path("..", "..", "..", "..", "..", "..", "..")
ruta_bulk_rna_seq_pipeline_absoluta <- normalizePath(ruta_bulk_rna_seq_pipeline)
nombre_proyecto <- basename(params$project_path)
nombre_experimento <- basename(params$experiment_name)


# Rutas de resultados Nextflow
ruta_nextflow_results <- file.path(ruta_bulk_rna_seq_pipeline, "resources", "02-nextflow-results")
ruta_proyecto_relativa <- file.path(ruta_bulk_rna_seq_pipeline, "resources", "02-nextflow-results", "01-project-data", nombre_proyecto)


# Construir ruta usando file.path
ruta_his_ReadCount <- file.path(params$project_path, "Analisis", nombre_experimento, "Readcount_results", "his-ReadCount.tab")
ruta_his_ReadCount_relativa <- file.path(ruta_proyecto_relativa, "Analisis", nombre_experimento, "Readcount_results", "his-ReadCount.tab")


# Construir ruta usando file.path
ruta_his_Size <- file.path(params$project_path, "Analisis", nombre_experimento, "Readcount_results", "his-Size.tab")
ruta_his_Size_relativa <- file.path(ruta_proyecto_relativa, "Analisis", nombre_experimento, "Readcount_results", "his-Size.tab")


# Rutas para guardar los archivos creados
ruta_archives_tmp_analisis <- file.path(ruta_bulk_rna_seq_pipeline, "resources", "01-essential", "02-archives", "02-tmp", "tab3-analisis")

ruta_total_lecturas_muestra_txt <- file.path(ruta_archives_tmp_analisis, "total_lecturas_muestra.txt")
ruta_genes_expresados_muestra_txt <- file.path(ruta_archives_tmp_analisis, "genes_expresados_muestra.txt")
ruta_genes_mayor_expresion_total_txt <- file.path(ruta_archives_tmp_analisis, "genes_mayor_expresion_total.txt")
```















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  TÍTULO Y RESUMEN CONTEXTUAL -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<div class="informe-titulo-tab">
  <h1>Tab</h1>
  <h2>Comprehensive Bioinformatic Analysis</h2>
</div>

<div class="informe-titulo-seccion-unique">
  <h1>Section 4</h1>
  <h2>Gene Expression Quantification</h2>
</div>

<div class="flecha-abajo">
  ▼
</div>


<div class="informe-resumen">
  <h2>Summary</h2>

  <p>
    In this section, <strong>gene expression quantification</strong> is carried out, a fundamental step following the <em>alignment</em> of reads to the reference genome. 
    The main objective is to <strong>determine how many reads are assigned to each gene</strong>, generating a <em>count matrix</em> that represents transcriptional activity in each analyzed sample. This information forms the basis for subsequent analyses of <strong>differential expression</strong> and <strong>functional enrichment</strong>.
  </p>

  <p>
    Quantification has been performed using the <em>alignment files</em> (<code>BAM</code>) generated previously, applying genomic annotations to identify regions corresponding to the genes. The result is an unnormalized expression matrix, stored in the file <code><strong>his-ReadCount.tab</strong></code>, where each row represents a gene and each column represents a sample from the experiment.
  </p>

  <p>
    In parallel, the file <code><strong>his-Size.tab</strong></code> has been generated, containing the <strong>length of each gene</strong> in base pairs (bp). 
    This information is crucial for performing <strong>proper normalization</strong> of the data later, as it allows correction for the bias introduced by variable gene lengths.
  </p>

  <p>
    To evaluate the quality and structure of the dataset, various <strong>exploratory visualizations</strong> of the raw counts are included. 
    These include the <em>interactive visualization of the count matrix</em>, <strong>gene length analysis</strong>, and a set of charts and metrics that assess the <em>distribution</em>, <em>consistency</em>, and <em>homogeneity</em> among samples.
  </p>

  <p>
    Key indicators are also examined, such as the <strong>total reads assigned</strong> per sample, the <strong>number of expressed genes</strong>, the <em>most abundantly transcribed genes</em>, and the <strong>distribution of expression levels</strong> through <em>logarithmic boxplots</em>. 
    This preliminary analysis allows detection of <strong>unusual deviations</strong> or outlier samples before proceeding to normalization and statistical analyses.
  </p>
</div>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  TABLA DE CONTENIDOS -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<div style="height: 15px; background-color: transparent;"></div>

<div class="titulo-toc">
  Table of contents for this section
</div>

<div class="toc">
  <ul>
    <li>
      <a href="#tab3-seccion4">
        4<span>.</span> Gene Expression Quantification
      </a>
      <ul>
        <li>
          <a href="#tab3-seccion4.1">
            4.1. Visualization of the raw count matrix
          </a>
        </li>
        <li>
          <a href="#tab3-seccion4.2">
            4.2. Gene length visualization
          </a>
        </li>
        <li>
          <a href="#tab3-seccion4.3">
            4.3. Preliminary exploration of raw counts
          </a>
          <ul>
            <li>
              <a href="#tab3-seccion4.3.1">
                4.3.1. Total reads assigned per sample
              </a>
            </li>
            <li>
              <a href="#tab3-seccion4.3.2">
                4.3.2. Number of expressed genes per sample
              </a>
            </li>
            <li>
              <a href="#tab3-seccion4.3.3">
                4.3.3. Genes with highest total expression
              </a>
            </li>
            <li>
              <a href="#tab3-seccion4.3.4">
                4.3.4. Distribution of counts per sample (log10 boxplot)
              </a>
            </li>
          </ul>
        </li>
      </ul>
    </li>
  </ul>
</div>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  REDACCIÓN -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<div style="height: 25px; background-color: transparent;"></div>

<div class="titulo1" id="tab3-seccion4">
  4<span>.</span> Gene Expression Quantification
</div>

<p>
  After aligning the reads to the reference genome, the next fundamental step in an RNA-Seq analysis is to
  quantify gene expression. This process allows determining how many reads are associated with each gene, providing an estimate
  of the transcriptional activity level in each sample.
</p>

<p>
  In this project, quantification has been performed by counting reads aligned to annotated gene regions, generating
  a raw (unnormalized) expression matrix. This count was obtained from the <code>SAM</code>/<code>BAM</code> files generated during alignment.
</p>

<p>
  For this purpose, two fundamental files have been produced:
</p>

<ul>
  <li>
    <strong><code>his-ReadCount.tab</code></strong> <strong>&rArr;</strong> contains the raw count matrix, where rows represent genes and columns
    correspond to the different samples of the experiment. The values indicate the number of reads that align to each gene in each sample.
  </li>
  <li>
    <strong><code>his-Size.tab</code></strong> <strong>&rArr;</strong> stores the length (in base pairs) of each gene. This information is essential for normalizing
    expression counts and for comparing expression levels between genes of different lengths or between samples with varying sequencing depth.
  </li>
</ul>

<p>
  The combination of these two files allows generating normalized expression metrics such as <code>RPKM</code> or <code>TPM</code>, which will be analyzed in a later section.
</p>











<div style="height: 20px; background-color: transparent;"></div>

<div class="titulo2" id="tab3-seccion4.1">
  4.1. Visualization of the raw count matrix
</div>

<p>
  Below is an interactive view of the unnormalized gene expression matrix, extracted from the file <code>his-ReadCount.tab</code>. 
  This matrix reflects the number of reads aligned to each gene in each of the analyzed samples and serves as the starting point for subsequent 
  normalization analyses and detection of differentially expressed genes.
</p>

<p>
  In the table shown below, it is possible to <strong>search directly by gene identifier</strong> using the integrated search bar. 
  Additionally, by <strong>clicking on the column headers</strong>, values can be sorted in ascending or descending order, facilitating 
  detailed exploration of expression levels per sample.
</p>

```{r}
library(DT)

# Leer el archivo y mostrar tabla interactiva
his_ReadCount <- read.delim(ruta_his_ReadCount, row.names = 1)

# Mostrar tabla interactiva con DT
DT::datatable(
  his_ReadCount,
  options = list(
    scrollX = TRUE,
    scrollY = "500px",
    scrollCollapse = TRUE,
    paging = FALSE,
    lengthChange = FALSE,
    info = FALSE,
    dom = 'ft', # Solo tabla
    language = list(
      url = '//cdn.datatables.net/plug-ins/1.13.4/i18n/es-ES.json'
    )
  ),
  class = 'stripe hover compact',
)
```

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center; margin-top: 5px;">
  <a href="{ruta_his_ReadCount_relativa}" 
    download="{basename(ruta_his_ReadCount_relativa)}" 
    class="boton-file-download">
    <i class="fa-solid fa-file-arrow-down"></i> Download file "{basename(ruta_his_ReadCount_relativa)}"
  </a>
</p>
'), sep = "\n")
```

<p>
  If you want to view the complete file directly, you can do so using the embedded viewer below or open it in a new tab:
</p>

```{r, results='asis'}
cat(glue::glue('
<div class="iframe-container">
  <iframe src="{ruta_his_ReadCount_relativa}" frameborder="0" style="height: 500px;"></iframe>
</div>

<p style="text-align:center;">
  <a href="{ruta_his_ReadCount_relativa}"
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-file-lines"></i> Open file in a new tab
  </a>
</p>
'), sep = "\n")
```










<div style="height: 20px; background-color: transparent;"></div>

<div class="titulo2" id="tab3-seccion4.2">
  4.2. Gene length visualization
</div>

<p>
  In addition to expression counts, it is necessary to consider the length of each gene to apply proper normalization methods, such as <code>RPKM</code> or <code>TPM</code>. 
  Gene length allows correction for the bias introduced by the fact that longer genes tend to accumulate more reads simply due to their size, 
  regardless of their actual expression level.
</p>

<p>
  Gene length information is contained in the file <code>his-Size.tab</code>, which provides the measurement in base pairs (bp) 
  for each transcript or gene included in the analysis. This file was generated using the genomic annotations corresponding to the reference genome used 
  during the alignment phase.
</p>

<p>
  Below is an interactive table with the gene length data. This table allows <strong>sorting lengths in ascending or descending order</strong> 
  by clicking on the column header, and <strong>searching for specific genes by their identifier</strong> using the available search bar. 
  This visualization facilitates the inspection of potential biases in the length distribution that could influence subsequent analyses.
</p>

```{r}
# Leer el archivo y mostrar tabla interactiva
his_Size <- read.delim(ruta_his_Size, row.names = 1)

# Mostrar tabla interactiva con DT
DT::datatable(
  his_Size,
  options = list(
    scrollX = TRUE,
    scrollY = "500px",
    scrollCollapse = TRUE,
    paging = FALSE,
    lengthChange = FALSE,
    info = FALSE,
    dom = 'ft', # Solo tabla
    language = list(
      url = '//cdn.datatables.net/plug-ins/1.13.4/i18n/es-ES.json'
    )
  ),
  class = 'stripe hover compact',
)
```

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center; margin-top: 5px;">
  <a href="{ruta_his_Size_relativa}" 
    download="{basename(ruta_his_Size_relativa)}" 
    class="boton-file-download">
    <i class="fa-solid fa-file-arrow-down"></i> Download file "{basename(ruta_his_Size_relativa)}"
  </a>
</p>
'), sep = "\n")
```

<p>
  If you want to view the complete file directly, you can do so using the embedded viewer below or open it in a new tab:
</p>

```{r, echo=FALSE, results='asis'}
cat(glue::glue('
<div class="iframe-container">
  <iframe src="{ruta_his_Size_relativa}" frameborder="0" style="height: 500px;"></iframe>
</div>

<p style="text-align:center;">
  <a href="{ruta_his_Size_relativa}"
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-file-lines"></i> Open file in a new tab
  </a>
</p>
'), sep = "\n")
```










<div style="height: 20px; background-color: transparent;"></div>

<div class="titulo2" id="tab3-seccion4.3">
  4.3. Preliminary exploration of raw counts
</div>

<p>
  Before applying normalization techniques and statistical analysis, it is essential to perform a preliminary inspection of the obtained expression counts.
  This exploration aims to identify possible anomalies, assess the overall distribution of the data, and check consistency among samples.
</p>

<p>
  Using the file <code>his-ReadCount.tab</code>, which contains the unnormalized expression matrix, various exploratory analyses are performed that allow:
</p>

<ul>
  <li>Calculating the total number of reads assigned to each sample.</li>
  <li>Determining the number of expressed genes per sample.</li>
  <li>Identifying the genes with the highest expression levels.</li>
  <li>Visualizing the overall distribution of counts per gene and per sample.</li>
</ul>

<p>
  These metrics and visualizations provide a first impression of the dataset quality and allow anticipating potential sources of variability that could affect differential expression analysis results.
</p>








<div style="height: 15px; background-color: transparent;"></div>

<div class="titulo3" id="tab3-seccion4.3.1">
  4.3.1. Total reads assigned per sample
</div>

<p>
  To begin the exploration of the counts, the total number of reads assigned to each sample is calculated. 
  This value corresponds to the sum of all counts per sample and provides an initial estimate of the sequencing depth obtained in the experiment.
</p>

<p>
  Below is the resulting table, showing the total reads per sample, which can be explored interactively.
</p>

```{r}
library(DT)
library(ggplot2)
library(plotly)
library(dplyr)
library(stringr)
library(scales)
library(tidyverse)
library(tidyr)
library(tibble)

# Calcular número total de lecturas asignadas por muestra
total_recuentos_por_muestra <- colSums(his_ReadCount)
total_recuentos_por_muestra_df <- data.frame(
  Muestra = names(total_recuentos_por_muestra),
  TotalRecuentos = as.integer(total_recuentos_por_muestra)
)

# Convertir a data frame para ggplot2
df_total <- data.frame(
  muestra = names(total_recuentos_por_muestra),
  total_lecturas = as.integer(total_recuentos_por_muestra)
)

# Asegurarnos que 'targets' está cargado previamente
targets_experimento <- paste0("targets_", nombre_experimento, ".txt")
ruta_targets_experimento <- file.path(params$project_path, "Resultados", targets_experimento)
ruta_targets_experimento_relativa <- file.path(ruta_proyecto_relativa, "Resultados", targets_experimento)
targets <- read.table(ruta_targets_experimento, header = TRUE, fill = TRUE, stringsAsFactors = FALSE)

# Limpiar y unir condición
df_total$muestra <- str_remove(df_total$muestra, "_nat$")
df_total <- df_total %>%
  left_join(targets %>% select(Filename, Type), by = c("muestra" = "Filename")) %>%
  rename(condicion = Type) %>%
  select(muestra, condicion, total_lecturas) %>%
  arrange(desc(total_lecturas))

# Guardar dataframe generado en formato .txt
write_delim(df_total, ruta_total_lecturas_muestra_txt, delim = "\t")

# Mostrar tabla interactiva con DT
DT::datatable(
  df_total,
  options = list(
    scrollX = TRUE,
    scrollY = "500px",
    scrollCollapse = TRUE,
    paging = FALSE,
    lengthChange = FALSE,
    info = FALSE,
    dom = 'ft', # Solo tabla
    language = list(
      url = '//cdn.datatables.net/plug-ins/1.13.4/i18n/es-ES.json'
    )
  ),
  class = 'stripe hover compact',
)
```

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center; margin-top: 5px;">
  <a href="{ruta_total_lecturas_muestra_txt}" 
    download="{basename(ruta_total_lecturas_muestra_txt)}" 
    class="boton-file-download">
    <i class="fa-solid fa-file-arrow-down"></i> Download file "{basename(ruta_total_lecturas_muestra_txt)}"
  </a>
</p>

<p>
  These results that we have just visualized have been saved in a file named <code>{basename(ruta_total_lecturas_muestra_txt)}</code>, which can be downloaded using the button provided above. Additionally, we can view this file using the following <code>iframe</code> or explore it in more detail in a new tab.
</p>

<div class="iframe-container">
  <iframe src="{ruta_total_lecturas_muestra_txt}" frameborder="0" style="height: 500px;"></iframe>
</div>

<p style="text-align:center;">
  <a href="{ruta_total_lecturas_muestra_txt}" 
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-file-lines"></i> Open file in a new tab
  </a>
</p>
'), sep = "\n")
```

<p>
  Finally, the following interactive bar plot is provided, which allows comparison of sequencing depth between conditions and replicates, providing a more dynamic and visual way to observe these results.
</p>

<div class="plot-center">

```{r}
# Gráfico de número total de lecturas por muestra con plotly
p <- plot_ly(
  data = df_total,
  x = ~muestra,
  y = ~total_lecturas,
  type = "bar",
  color = ~condicion,
  colors = "Set2",
  text = ~comma(total_lecturas),
  textposition = "outside",
  hoverinfo = "text",
  hovertext = ~paste0("Sample: ", muestra, "<br>Reads: ", comma(total_lecturas))
) %>%
  layout(
    title = "Total reads per sample",
    xaxis = list(title = "Sample", tickangle = -45, tickfont = list(size = 10)),
    yaxis = list(
      title = "Number of reads",
      range = c(0, max(df_total$total_lecturas) * 1.15)
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.5,
      xanchor = "center"
    ),
    margin = list(t = 100, b = 170),
    autosize = TRUE
  )

# Representar el gráfico
p

# Ruta para guardar el gráfico
ruta_graph_total_lecturas_muestra_html <- file.path(ruta_archives_tmp_analisis, "graph_total_lecturas_muestra.html")

# Guardar con título personalizado en la pestaña del navegador
saveWidget(
  widget = p,
  file = ruta_graph_total_lecturas_muestra_html,
  selfcontained = TRUE,
  title = "Total reads per sample"
)
```

</div>

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center;">
  <a href="{ruta_graph_total_lecturas_muestra_html}" 
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-chart-column"></i> Open graph in full screen
  </a>
</p>
'), sep = "\n")
```








<div style="height: 15px; background-color: transparent;"></div>

<div class="titulo3" id="tab3-seccion4.3.2">
  4.3.2. Number of expressed genes per sample
</div>

<p>
  After analyzing sequencing depth, the next step is to evaluate how many genes were detected as expressed in each sample.
  For this, a gene is considered expressed if it has a count greater than zero in a given sample. This analysis helps identify potential low-quality samples or technical issues, as an unusually low number of expressed genes could indicate poor RNA capture or amplification efficiency.
</p>

<p>
  Below is the resulting table, showing the number of expressed genes per sample, which can be explored interactively.
</p>

```{r}
# Calcular número de genes expresados por muestra (genes con recuentos > 0)
genes_expresados_por_muestra <- colSums(his_ReadCount > 0)
genes_expresados_por_muestra_df <- data.frame(
  Muestra = names(genes_expresados_por_muestra),
  GenesExpresados = as.integer(genes_expresados_por_muestra)
)

# Crear data frame
df_genes_exp <- data.frame(
  muestra = names(genes_expresados_por_muestra),
  genes_expresados = as.integer(genes_expresados_por_muestra)
)

# Añadir la columna 'condicion' uniendo por el nombre de muestra
df_genes_exp$muestra <- str_remove(df_genes_exp$muestra, "_nat$")
df_genes_exp <- df_genes_exp %>%
  left_join(targets %>% select(Filename, Type), by = c("muestra" = "Filename")) %>%
  rename(condicion = Type) %>%
  select(muestra, condicion, genes_expresados) %>%
  arrange(desc(genes_expresados))

# Guardar dataframe generado en formato .txt
write_delim(df_genes_exp, ruta_genes_expresados_muestra_txt, delim = "\t")

# Mostrar tabla interactiva con DT
DT::datatable(
  df_genes_exp,
  options = list(
    scrollX = TRUE,
    scrollY = "500px",
    scrollCollapse = TRUE,
    paging = FALSE,
    lengthChange = FALSE,
    info = FALSE,
    dom = 'ft', # Solo tabla
    language = list(
      url = '//cdn.datatables.net/plug-ins/1.13.4/i18n/es-ES.json'
    )
  ),
  class = 'stripe hover compact',
)
```

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center; margin-top: 5px;">
  <a href="{ruta_genes_expresados_muestra_txt}" 
    download="{basename(ruta_genes_expresados_muestra_txt)}" 
    class="boton-file-download">
    <i class="fa-solid fa-file-arrow-down"></i> Download file "{basename(ruta_genes_expresados_muestra_txt)}"
  </a>
</p>

<p>
  These results that we have just visualized have been saved in a file named <code>{basename(ruta_genes_expresados_muestra_txt)}</code>, which can be downloaded using the button provided above. Additionally, we can view this file using the following <code>iframe</code> or explore it in more detail in a new tab.
</p>

<div class="iframe-container">
  <iframe src="{ruta_genes_expresados_muestra_txt}" frameborder="0" style="height: 500px;"></iframe>
</div>

<p style="text-align:center;">
  <a href="{ruta_genes_expresados_muestra_txt}" 
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-file-lines"></i> Open file in a new tab
  </a>
</p>
'), sep = "\n")
```

<p>
  Finally, the following interactive bar plot is provided, which facilitates comparison between experimental conditions and provides a more dynamic and visual way to observe these results.
</p>

<div class="plot-center">

```{r}
# Gráfico: genes expresados por muestra (>0) con plotly
p <- plot_ly(
  data = df_genes_exp,
  x = ~muestra,
  y = ~genes_expresados,
  type = "bar",
  color = ~condicion,
  colors = "Set2",  # mismo estilo que ggplot (Set1)
  text = ~comma(genes_expresados),
  textposition = "outside",
  hoverinfo = "text",
  hovertext = ~paste0("Sample: ", muestra, "<br>Expressed genes: ", comma(genes_expresados))
) %>%
  layout(
    title = "Number of expressed genes per sample",
    xaxis = list(
      title = "Sample",
      tickangle = -45,
      tickfont = list(size = 10)
    ),
    yaxis = list(
      title = "Genes with counts > 0",
      range = c(0, max(df_genes_exp$genes_expresados) * 1.15),
      separatethousands = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.52,
      xanchor = "center"
    ),
    margin = list(t = 100, b = 170),
    autosize = TRUE
  )

# Representar el gráfico
p

# Ruta para guardar el gráfico
ruta_graph_genes_expresados_muestra_html <- file.path(ruta_archives_tmp_analisis, "graph_genes_expresados_muestra.html")

# Guardar con título personalizado en la pestaña del navegador
saveWidget(
  widget = p,
  file = ruta_graph_genes_expresados_muestra_html,
  selfcontained = TRUE,
  title = "Number of expressed genes per sample"
)
```

</div>

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center;">
  <a href="{ruta_graph_genes_expresados_muestra_html}" 
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-chart-column"></i> Open graph in full screen
  </a>
</p>
'), sep = "\n")
```








<div style="height: 15px; background-color: transparent;"></div>

<div class="titulo3" id="tab3-seccion4.3.3">
  4.3.3. Genes with highest total expression
</div>

<p>
  To identify the genes with the highest transcriptional activity within the dataset, the total sum of counts per gene across all samples is calculated.
  This metric allows detection of genes that show the highest expression levels globally, which could play a relevant role in the biological context of the experiment.
</p>

<p>
  Below is the resulting table, with genes ranked according to their highest cumulative expression, in descending order of total reads.
</p>

```{r}
# Obtener los genes con mayor nivel de expresión total (suma por fila)
top_genes<- rowSums(his_ReadCount) |>
  sort(decreasing = TRUE)

top_genes_df <- data.frame(
  Gen = names(top_genes),
  TotalRecuentos = as.integer(top_genes)
)

top_genes_20 <- rowSums(his_ReadCount) |>
  sort(decreasing = TRUE) |>
  head(20)

top_genes_20_df <- data.frame(
  Gen = names(top_genes_20),
  TotalRecuentos = as.integer(top_genes_20)
)

# Guardar dataframe generado en formato .txt
write_delim(top_genes_df, ruta_genes_mayor_expresion_total_txt, delim = "\t")

# Crear la tabla interactiva
DT::datatable(
  top_genes_df,
  options = list(
    scrollX = TRUE,
    scrollY = "500px",
    scrollCollapse = TRUE,
    paging = FALSE,
    lengthChange = FALSE,
    info = FALSE,
    dom = 'ft', # Solo tabla
    language = list(
      url = '//cdn.datatables.net/plug-ins/1.13.4/i18n/es-ES.json'
    )
  ),
  class = 'stripe hover compact',
)
```

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center; margin-top: 5px;">
  <a href="{ruta_genes_mayor_expresion_total_txt}" 
    download="{basename(ruta_genes_mayor_expresion_total_txt)}" 
    class="boton-file-download">
    <i class="fa-solid fa-file-arrow-down"></i> Download file "{basename(ruta_genes_mayor_expresion_total_txt)}"
  </a>
</p>

<p>
  These results that we have just visualized have been saved in a file named <code>{basename(ruta_genes_mayor_expresion_total_txt)}</code>, which can be downloaded using the button provided above. Additionally, we can view this file using the following <code>iframe</code> or explore it in more detail in a new tab.
</p>

<div class="iframe-container">
  <iframe src="{ruta_genes_mayor_expresion_total_txt}" frameborder="0" style="height: 500px;"></iframe>
</div>

<p style="text-align:center;">
  <a href="{ruta_genes_mayor_expresion_total_txt}" 
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-file-lines"></i> Open file in a new tab
  </a>
</p>
'), sep = "\n")
```

<p>
  Finally, the following interactive bar plot is provided, showing the 20 genes with the highest expression, allowing assessment of their relative contribution within the dataset.
</p>

<div class="plot-center">

```{r}
# Ordenar los genes de forma descendente para el gráfico
top_genes_20_df$Gen <- factor(top_genes_20_df$Gen, levels = rev(top_genes_20_df$Gen))

# Gráfico: top 20 genes más expresados con plotly (horizontal)
p <- plot_ly(
  data = top_genes_20_df,
  x = ~TotalRecuentos,
  y = ~Gen,
  type = "bar",
  orientation = "h",
  marker = list(
    color = "#098",
    line = list(color = "black", width = 1)
  ),
  text = ~comma(TotalRecuentos),
  textposition = "outside",
  hoverinfo = "text",
  hovertext = ~paste0("Gene: ", Gen, "<br>Counts: ", comma(TotalRecuentos))
) %>%
  layout(
    title = "Top 20 most expressed genes",
    xaxis = list(
      title = "Total counts",
      range = c(0, max(top_genes_20_df$TotalRecuentos) * 1.2),
      separatethousands = TRUE
    ),
    yaxis = list(
      title = "Gene",
      automargin = TRUE,
      tickfont = list(size = 10)
    ),
    margin = list(l = 100, t = 80, r = 40),
    autosize = TRUE
  )

# Representar el gráfico
p

# Ruta para guardar el gráfico
ruta_graph_genes_mayor_expresion_total_html <- file.path(ruta_archives_tmp_analisis, "graph_genes_mayor_expresion_total.html")

# Guardar con título personalizado en la pestaña del navegador
saveWidget(
  widget = p,
  file = ruta_graph_genes_mayor_expresion_total_html,
  selfcontained = TRUE,
  title = "Top 20 most expressed genes"
)
```

</div>

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center;">
  <a href="{ruta_graph_genes_mayor_expresion_total_html}" 
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-chart-column"></i> Open graph in full screen
  </a>
</p>
'), sep = "\n")
```








<div style="height: 15px; background-color: transparent;"></div>

<div class="titulo3" id="tab3-seccion4.3.4">
  4.3.4. Distribution of counts per sample (log10 boxplot)
</div>

<p>
  Finally, to evaluate variability and the distribution of expression counts at the genomic level in each sample, a boxplot is generated using a logarithmic transformation (log10).
  This representation allows identification of potential outliers, differences in data dispersion, and assessment of homogeneity among replicates and experimental conditions.
</p>

<p>
  Zero values are excluded to avoid issues with the logarithmic transformation, and an interactive visualization is used to facilitate detailed exploration of the distribution in each sample.
</p>


<div style="height: 15px; background-color: transparent;"></div>

```{r}
# Convertir a formato largo
his_ReadCount_long <- his_ReadCount |>
  rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "muestra", values_to = "recuento")

# Filtrar recuentos mayores que cero antes de graficar
his_ReadCount_long_filtrado <- his_ReadCount_long |>
  filter(recuento > 0)

# Hacer el left join con targets para añadir la columna 'Type' como 'condicion'
his_ReadCount_long_filtrado <- his_ReadCount_long_filtrado %>%
  mutate(muestra_clean = str_remove(muestra, "_nat$")) %>%
  select(-muestra) %>%
  rename(muestra = muestra_clean) %>%
  left_join(targets %>% select(Filename, Type), by = c("muestra" = "Filename")) %>%
  rename(condicion = Type)
```

<div class="plot-center">

```{r}
# Gráfico: distribución logarítmica de recuentos por muestra (boxplot)
p <- plot_ly(
  data = his_ReadCount_long_filtrado,
  x = ~muestra,
  y = ~recuento,
  color = ~condicion,
  colors = "Set2",
  type = "box",
  boxpoints = "outliers",
  marker = list(size = 3, opacity = 0.5, line = list(width = 0)),
  line = list(color = "black"),
  opacity = 0.8
) %>%
  layout(
    title = list(
      text = "Logarithmic distribution of counts per sample<br><sub>Each boxplot represents gene expression levels (log10) per sample</sub>",
      y = 0.95
    ),
    xaxis = list(
      title = "Sample",
      tickangle = -45,
      tickfont = list(size = 10)
    ),
    yaxis = list(
      title = "Count (log10)",
      type = "log",
      separatethousands = TRUE
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.52,
      xanchor = "center"
    ),
    margin = list(t = 100, b = 170),
    autosize = TRUE
  )

# Representar el gráfico
p

# Ruta para guardar el gráfico
ruta_graph_distribucion_expresion_muestra_html <- file.path(ruta_archives_tmp_analisis, "graph_distribucion_expresion_muestra.html")

# Guardar con título personalizado en la pestaña del navegador
saveWidget(
  widget = p,
  file = ruta_graph_distribucion_expresion_muestra_html,
  selfcontained = TRUE,
  title = "Expression distribution per sample"
)
```

</div>

```{r, results='asis'}
cat(glue::glue('
<p style="text-align:center;">
  <a href="{ruta_graph_distribucion_expresion_muestra_html}" 
    target="_blank" 
    class="boton-iframe">
    <i class="fa-solid fa-chart-column"></i> Open graph in full screen
  </a>
</p>
'), sep = "\n")
```










<div class="divider"></div>

<p>
  After this <strong>preliminary exploration of gene expression counts</strong>, in which key aspects such as total reads assigned, the distribution of expressed genes, and overall expression per sample have been examined, the next step will be to apply a <em>normalization process</em> to allow fair comparison between samples and reduce potential technical biases. Next, a <a href="05.0-analisis-estadistico.html#" class="normalink">statistical analysis of gene expression</a> will be conducted, which will include <strong>post-normalization quality assessment</strong>, identification of <strong>differentially expressed genes</strong>, and performance of a <strong>functional and enrichment analysis</strong> to interpret the results in the corresponding biological context.
</p>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  FOOTER -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<footer>
  <a href="#" class="arriba"><i class="fa-solid fa-angles-up"></i></a>
</footer>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  CARGAR SCRIPTS DE JAVASCRIPT -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<script type="module" src="../../../../04-javascript/chatbot.js"></script>
<script type="module" src="../../../../04-javascript/script.js"></script>
