---
title: "IPBLN Bioinformatics Report"

include-in-header:
  text: |
    <link rel="shortcut icon" href="../../../../../01-images/icons/favicon.svg"/>
    <link rel="icon" type="image/x-icon" href="../../../../../01-images/icons/favicon.ico">

params:
  project_path: ""
  experiment_name: ""
  report_language: ""
  report_version: ""
---


<!-- &&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&& -->
<!--  TAB 3: ANÁLISIS | 5.2. EVALUACIÓN DEL CONTROL DE CALIDAD TRAS LA NORMALIZACIÓN -->
<!-- &&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&&& -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  CHATBOT -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

```{=html}
<!-- Contenedor del chat -->
<div id="chat-container">
  <button id="chat-toggle"></button>
  <div id="chat-box">
    <div id="chat-info" class="chat-info">
      <h1>
        Mini Chat RAG (beta)
      </h1>

      <p>
        Hello! I am <strong>Geni</strong>, the intelligent assistant of <strong>GenoScribe</strong>.  
        I am here to help you interactively explore the content of this bioinformatics report.
      </p>

      <p>
        When you ask me a question, I first try to recognize whether it matches any of the patterns or expressions I know.  
        If I find a match, I will respond directly with a predefined answer, designed to be fast, clear, and even a bit witty.  
        If I do not recognize the pattern, I then activate my search tools: I generate vector representations (embeddings) and look for the most relevant fragments across multiple documents—including the report itself, external PDF and HTML files, and question-and-answer (QA) sessions.  
        Based on that information, I create a summary that aims to provide you with a coherent and useful response grounded in the existing content.
      </p>

      <p>
        Please note that this environment is experimental. I do not use large language models, so some responses may be approximate or incomplete.  
        The main goal is to provide a fast, understandable, and reproducible way to visualize the information contained in the documents, enabling a more dynamic exploration of the report.
      </p>

      <p>
        Currently, results may vary in accuracy, as I use lightweight local models to ensure the application runs in any environment without the need for external servers.  
        However, the system structure is designed to significantly improve its performance in the future through integration with more advanced models or external APIs.  
        To get started, simply type your question in the field below and let me take care of the rest. I promise to put all my code into it!
      </p>
    </div>
    <div id="chat-log"></div>
    <input id="chat-input" type="text" placeholder="Escribe tu pregunta..."/>
    <button id="chat-send" title="Enviar">
      <svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 24 24">
        <path d="M2 21l21-9L2 3v7l15 2-15 2v7z"/>
      </svg>
    </button>
  </div>
</div>
```















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  IMPORTACIONES -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<!-- Incluir Font Awesome -->
<link href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/6.5.0/css/all.min.css" rel="stylesheet">















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  CÓDIGO GLOBAL -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

```{r}
library(stringr)
library(glue)
library(fs)
library(dplyr)

# Rutas clave
ruta_qmd <- getwd()
ruta_bulk_rna_seq_pipeline <- file.path("..", "..", "..", "..", "..", "..", "..")
ruta_bulk_rna_seq_pipeline_absoluta <- normalizePath(ruta_bulk_rna_seq_pipeline)
nombre_proyecto <- basename(params$project_path)
nombre_experimento <- basename(params$experiment_name)


# Rutas de resultados Nextflow
ruta_nextflow_results <- file.path(ruta_bulk_rna_seq_pipeline, "resources", "02-nextflow-results")
ruta_proyecto_relativa <- file.path(ruta_bulk_rna_seq_pipeline, "resources", "02-nextflow-results", "01-project-data", nombre_proyecto)


# Rutas de informe de control de calidad tras la normalización
ruta_experimento_QC <- file.path(params$project_path, "Resultados", nombre_experimento, paste0(nombre_experimento, "_QC.pdf"))
ruta_experimento_QC_relativa <- file.path(ruta_proyecto_relativa, "Resultados", nombre_experimento, paste0(nombre_experimento, "_QC.pdf"))
```















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  TÍTULO Y RESUMEN CONTEXTUAL -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<div class="informe-titulo-tab">
  <h1>Tab</h1>
  <h2>Comprehensive Bioinformatics Analysis</h2>
</div>

<div class="informe-titulo-seccion">
  <h1>Section 5</h1>
  <h2>Statistical Analysis of Gene Expression</h2>
</div>

<div class="informe-titulo-subseccion-unique">
  <h1>Subsection 5.2</h1>
  <h2>Post-Normalization Quality Control Assessment</h2>
</div>

<div class="flecha-abajo">
  ▼
</div>


<div class="informe-resumen">
  <h2>Summary</h2>

  <p>
    This section provides a <strong>comprehensive assessment of quality control</strong> after the normalization of gene expression data. This step is crucial to verify that normalization has properly corrected <em>technical variations</em> inherent to the experimental process without affecting the <em>true biological structure</em> of the analyzed samples.
  </p>

  <p>
    To achieve this, various exploratory techniques and statistical visualizations are applied to assess the homogeneity and consistency of normalized data. These techniques include <strong>Principal Component Analysis (PCA)</strong>, which facilitates the identification of clustering and potential outliers; the calculation of the <strong>correlation matrix</strong> between samples, which quantifies overall similarities; and <strong>hierarchical clustering</strong>, which organizes samples according to shared expression patterns.
  </p>

  <p>
    Additionally, <strong>heatmaps</strong> and distribution plots such as <em>boxplots</em> and <em>density plots</em> are generated, providing a comprehensive view of the variability and consistency of gene expression after adjustment. These tools help detect potential problematic samples or residual artifacts that could affect biological interpretation.
  </p>

  <p>
    All these analyses and visualizations are compiled in a <strong>consolidated PDF report</strong>, available for consultation and download at the specified path, serving as a valuable guide to validate dataset quality and ensure the reliability of subsequent differential expression and functional enrichment analyses.
  </p>
</div>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  TABLA DE CONTENIDOS -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<div style="height: 15px; background-color: transparent;"></div>

<div class="titulo-toc">
  Table of contents of this subsection
</div>

<div class="toc">
  <ul>
    <li>
      <a href="05-Analisis-estadistico.html#tab3-seccion5">
        5<span>.</span> Statistical analysis of gene expression
      </a>
      <ul>
        <li>
          <a href="#tab3-seccion5.2">
            5.2. Evaluation of quality control after normalization
          </a>
        </li>
      </ul>
    </li>
  </ul>
</div>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  REDACCIÓN -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<div style="height: 20px; background-color: transparent;"></div>

<div class="titulo2" id="tab3-seccion5.2">
  5.2. Evaluation of quality control after normalization
</div>

<p>
  After applying the <strong>normalization process</strong> to the gene expression matrix, it is essential to perform a <strong>comprehensive assessment of data quality</strong>.
  This step allows verifying that the <strong>preprocessing</strong> has properly corrected the <em>sources of technical variation</em> without altering the <em>biological structure</em> of the samples.
</p>

<p>
  For this purpose, a series of <strong>diagnostic plots</strong> have been generated from the normalized matrix, using exploratory approaches such as:
</p>

<ul>
  <li><strong>Principal Component Analysis (PCA)</strong></li>
  <li><strong>Sample correlation matrix</strong></li>
  <li><strong>Hierarchical clustering</strong></li>
  <li><strong>Heatmaps</strong></li>
  <li><strong>Expression distributions (boxplots, density plots)</strong></li>
</ul>

<p>
  This set of visualizations has been compiled into a single <strong>PDF report</strong>, available, at the time of generating this report, in the following directory:
</p>

```{r, echo=FALSE, results="asis"}
cat(glue::glue('
  <code>{ruta_experimento_QC}</code>
'), sep = "\n")
```

<p>
  And it can be viewed in the following interactive window:
</p>

```{r, results='asis'}
if (file.exists(ruta_experimento_QC_relativa)) {
  cat(glue::glue('
  <div class="iframe-container">
    <iframe src="{ruta_experimento_QC_relativa}" frameborder="0" style="height: 700px;"></iframe>
  </div>

  <div style="display: flex; justify-content: center; gap: 25px;">
    <a href="{ruta_experimento_QC_relativa}"
      target="_blank" 
      class="boton-iframe">
      <i class="fa-solid fa-file-lines"></i> Open report in a new tab
    </a>

    <a href="{ruta_experimento_QC_relativa}"
      download="{basename(ruta_experimento_QC_relativa)}"
      class="boton-iframe-download">
      <i class="fa-solid fa-file-arrow-down"></i> Download report
    </a>
  </div>
  '), sep = "\n")

} else {
  cat(glue::glue('
    <div class="alert-warning alert-box">
      <i class="fa-solid fa-triangle-exclamation"></i>
      <b>Warning</b><br>
      File "{basename(ruta_experimento_QC)}" not found in the specified path:<br>
      <code>{ruta_experimento_QC}</code>.<br>
      Please ensure this file is correctly defined in this directory.
    </div>
  '), sep = "\n")
}
```

<p>
  More specifically, this document contains the following types of plots:
</p>

<ul>
  <li>
    <strong>Log-CPM density curves</strong> <strong>&rArr;</strong> allow visualizing the distribution of expression values (on a logarithmic scale) for each sample before and after low-expression filtering. These curves help detect differences between samples and assess homogeneity after filtering.
  </li>
  <li>
    <strong>Expression boxplots per sample</strong> <strong>&rArr;</strong> show the statistical distribution of normalized values for each sample, before and after normalization. They help identify potential outliers or systematic biases that have been removed.
  </li>
  <li>
    <strong>Library sizes</strong> <strong>&rArr;</strong> indicate the total number of reads assigned per sample, providing a measure of sequencing depth and justifying the need for normalization.
  </li>
  <li>
    <strong>Sample correlation matrix</strong> <strong>&rArr;</strong> shows the correlation coefficient between all pairs of samples based on their normalized gene expression profiles. 
    High correlation among biological replicates indicates experimental consistency and good data quality. Discrepancies may suggest outliers or technical errors.
  </li>
  <li>
    <strong>Principal Component Analysis (PCA) and MDS</strong> <strong>&rArr;</strong> provide two-dimensional representations of variance between samples, grouping them according to their gene expression profiles. Clear separation between experimental groups supports the presence of significant biological differences.
  </li>
  <li>
    <strong>Heatmap with hierarchical clustering</strong> <strong>&rArr;</strong> represents the most variable genes in the experiment, showing their expression across all samples. The resulting clustering allows evaluation of consistency between replicates and experimental conditions.
  </li>
</ul>

<p>
  These graphical representations help confirm the quality of the normalized data and provide a global view that facilitates interpretation of the biological patterns present in the experiment. 
  From this point, the analysis focuses on the identification of differentially expressed genes between conditions.
</p>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  FOOTER -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<footer>
  <a href="#" class="arriba"><i class="fa-solid fa-angles-up"></i></a>
</footer>















<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!--  CARGAR SCRIPTS DE JAVASCRIPT -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

<script type="module" src="../../../../04-javascript/chatbot.js"></script>
<script type="module" src="../../../../04-javascript/script.js"></script>
